Vagueness and imprecise imitation in signaling games

### 2016-02-28 paper/version_05/paper.tex
Tiny fixes.

We believe that vagueness in language and thought may have many reasons, not just one. We focus
here on one \emph{a priori} plausible reason for why vagueness is natural and pervasive. The
idea is that vagueness is, at least in part, due to imprecision in the perception of similar
stimuli and imprecision in the realization of similar responses. On this view, vagueness in
language may be seen as a necessary sub-optimality due to limitations in another domain of
cognition, first and foremost perception. It could then be speculated that the whole system,
e.g., perception and language together, may be an almost optimal adaptation in a larger frame
of reference, e.g., if we take into account the metabolic costs for increased perceptual
accuracy. We will not engage in such speculation here. Rather we will explore the consequences
of perceptual limitations on processes of meaning evolution in a suitable formal framework. In
other words, in order to address Lipman's challenge seriously, and not just hand-wave it away
by appeal to the naturalness of vagueness, it needs to be spelled out how exactly confusability
of stimuli could come into play in a process of meaning evolution and how it could lead to
vague but by-and-large informative signal meaning. The formal model that this paper introduces
does exactly that. But it also does more. We find that confusability of stimuli can regularize
and systematize evolving meaning. This suggests a possibly advantageous side-effect that a
natural cause of vagueness may have and that might counter some of the disadvantages that
vagueness may have for communicative efficiency \citep[c.f.][]{OConnor2013:The-Evolution-o}.

@@

To illustrate, let us briefly consider a sim-max game with six equiprobable states and two
messages, its equilibria and what vagueness intuitively would look like \citep[see][for a more
thorough discussion]{Jager2007:The-Evolution-o,OConnor2013:The-Evolution-o}. We assume that
utilities are strictly decreasing with decreasing
similarity. Figure~\ref{fig:EquilibriaVagueness} shows three pairs of sender and receiver
strategies. States and acts are arranged according to their similarity: the closer they are to
each other, the more similar they are. The pair in Figure~\ref{fig:ineffNonEq} is not an
equilibrium, because the sender's non-convex use of signals is suboptimal given the receiver's
behavior. In contrast, Figure~\ref{fig:efficientEq}Z shows a maximally efficient
equilibrium. This is a partial pooling equilibrium in the sense that the speaker uses the same
message for several states. This would be inefficient if there were more messages, but it is
maximally efficient for the two-message case. So, while partial pooling may entail
inefficiency, and while partial pooling can hamper the evolution of maximally efficient signal
use
\citep[e.g.][]{Hutteger:2007_Evol_Indicatives_Imperatives,Pawlowitsch2008:Why-Evolution-d,HutteggerSkyrms2010:Evolutionary-Dy},
this is orthogonal to our concerns about vagueness. A regular and natural yet vague signal use
would look like in Figure~\ref{fig:vagueUse}. This is not an equilibrium, but it gets close, so
to speak. It shows smooth transitions across similar states at the boundaries of categories and
across acts around the most prototypical instances (as indicated by decreasing thickness of
arrows in Figure~\ref{fig:vagueUse}).

>>>>

We believe that vagueness in language and thought may have many reasons, not just one. We focus
here on one \emph{a priori} plausible reason for why vagueness is natural and pervasive. The
idea is that vagueness is, at least in part, due to imprecision in the perception of similar
stimuli and imprecision in the realization of similar responses. On this view, vagueness in
language may be seen as a necessary sub-optimality due to limitations in another domain of
cognition, first and foremost perception. It could then be speculated that the whole system,
perception and language together, may be an almost optimal adaptation in a larger frame
of reference, e.g., if we take into account the metabolic costs for increased perceptual
accuracy. We will not engage in such speculation here. Rather we will explore the consequences
of perceptual limitations on processes of meaning evolution in a suitable formal framework. In
other words, in order to address Lipman's challenge seriously, and not just hand-wave it away
by appeal to the naturalness of vagueness, it needs to be spelled out how exactly confusability
of stimuli could come into play in a process of meaning evolution and how it could lead to
vague but by-and-large informative signal meaning. The formal model that this paper introduces
does exactly that. But it also does more. We find that confusability of stimuli can regularize
and systematize evolving meaning. This suggests a possibly advantageous side-effect that a
natural cause of vagueness may have and that might compensate some of the disadvantages that
vagueness may have for communicative efficiency \citep[c.f.][]{OConnor2013:The-Evolution-o}.

@@

To illustrate, let us briefly consider a sim-max game with six equiprobable states and two
messages, its equilibria and what vagueness intuitively would look like \citep[see][for a more
thorough discussion]{Jager2007:The-Evolution-o,OConnor2013:The-Evolution-o}. We assume that
utilities are strictly decreasing with decreasing
similarity. Figure~\ref{fig:EquilibriaVagueness} shows three pairs of sender and receiver
strategies. States and acts are arranged according to their similarity: the closer they are to
each other, the more similar they are. The pair in Figure~\ref{fig:ineffNonEq} is not an
equilibrium, because the sender's non-convex use of signals is suboptimal given the receiver's
behavior. In contrast, Figure~\ref{fig:efficientEq} shows a maximally efficient
equilibrium. This is a partial pooling equilibrium in the sense that the speaker uses the same
message for several states. This would be inefficient if there were more messages, but it is
maximally efficient for the two-message case. So, while partial pooling may entail
inefficiency, and while partial pooling can hamper the evolution of maximally efficient signal
use
\citep[e.g.][]{Hutteger:2007_Evol_Indicatives_Imperatives,Pawlowitsch2008:Why-Evolution-d,HutteggerSkyrms2010:Evolutionary-Dy},
this is orthogonal to our concerns about vagueness. A regular and natural yet vague signal use
would look like the pair in Figure~\ref{fig:vagueUse}. This is not an equilibrium, but it gets close, so
to speak. It shows smooth transitions across similar states at the boundaries of categories and
across acts around the most prototypical instances (as indicated by decreasing thickness of
arrows in Figure~\ref{fig:vagueUse}).
